\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-07-16 metatheory.tex
Adding comments.
Save, rewriting meta-theory

% Motivation of using QIIT
We start with the formulation of the syntax, which is based on the
formalization of a predicative \textit{Martin-Lof Type Theory} (MLTT)
given by~\citet{coquand2018canonicity}. Thus we will omit a great deal
of details about MLTT here\YZ{What details are omitted?} but focus on the new facility we introduced.

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic a type theoretical framework (like Agda), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We base our formulation of the syntax in this type theory instead of a more conventional and informal representation, becase (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out.

@@

\subsection{Syntax}

>>>>

% Motivation of using QIIT
We start with the formulation of the syntax in the declarative style, which is based on the
formalization of a predicative \textit{Martin-Lof Type Theory} (MLTT)
given by~\citet{coquand2018canonicity}. 
% Thus we will omit a great deal of details about MLTT here
\YZ{What details are omitted?} 
We will mainly focus on the new facility that we introduced

Though phrased in Latex, we are following~\citet{altkap2016} to use as
our meta-logic an extensional type theoretical framework (with Agda-like syntax), 
together with \textit{Quotient Inductive Inductive Type} (QIIT). Our work can be seen as following the practice ``type theory should eat itself''\cite{dybjer1995internal, chapman2009type}. We use as the formulation the intrinsically-typed debruijn syntax in this type-theoretical QIIT instead of a more conventional and untyped representation, because (1) we hope our formulation can be precise on the mathematical object we are using (2) it is easier to validate a proof because proof checking then is simply type checking in type theory (3) when using QIIT to formulate syntax, the induction principle used in reasoning about syntax is itself justified (4) we hope our formulation can be closer to a real theorem prover so that the reader can have an idea about how mechanized formalization can be carried out (5) We can still use named-variables for intuition and example illustration (6) using QIIT as syntax encoding can avoid the reader from the concept of \textit{presupposition}~\citet{gratzer-sterling-birkedal:2019}---these \textit{presupposition} can be understood as \textit{implicit parameters} of the declaration of the judgemental rules in QIIT.

@@

\subsection{Syntax}\YZ{Rewrite syntax part into name-variable style instead of currently debruijn indices, and do that for all the models we introduced}

### 2022-07-16 syntax.tex
Save, rewriting meta-theory

We formalize our language as a deep embedding in a type theory with QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing} and thus a predicative MLTT, but for readability we
omit the universe levels.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the (pseudo) Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just as
a well-formed QIIT type is not necessarily inhabited. These well-formedness rules are direct translations of the \textit{well-typedness} of QIIT.

@@

Now we highlight some other special features that differ from a more
conventional setting: (1) This modern formulation of type theory does not
have an operational semantics but only equality.\footnote{Later we will prove
canonicity which can imply a certain level of computational ability of
our theory}
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
(2) Instead of using meta-level substitution, we use explicit
substitution and de Bruijn indices in our intrinsic formulation. This is
also called substitution calculus in the literature and is favoured due
to its closedness to categorical semantic.
(3) We are dealing with quotient data, so we are actually manipulating
equivalence classes of terms instead of one representative (term) from the equivalence class.

>>>>

\subsubsection{Warm Up}
\textit{Quotient Inductive Inductive Type}~(QIIT) is an extension of \textit{inductive-inductive type} (IIT)~\cite{nordvall2010inductive}. IIT can be roughly considered as several inductive type mutually defined, in a more expressible way than mutual-inductive type. QIIT is allowing to have \textit{quotient}, in other words, equality results as part of the constructors. For example, in \mintinline{Agda}{data N2: Set where O : N2 | S : N2 → N2 | B : S (S x) ≡ x }, "B" will be a valid (equality) constructor, and "N2" is a set of two elements in our extensional setting.

We highlight some special features of declarative style in QIIT that differ from a
conventional setting: 
(1) This declarative formulation of type theory does not
have an operational semantics but only judgemental equality---$\beta$-reduction is now $\beta$-equality.\footnote{Later we will prove
canonicity which can imply a certain level of computational ability of
our theory} 
(2) We directly use the equality from QIIT to represent judgemental equality in our syntax, relieve from us the burden to explicit mentioning equivalence closure; however, this also mean that, \textit{every piece of the syntax data}, including all kinds of judgements (QIIT type) and their derivations will have equality information upon them  
%We will recover canonicity afterwards, which implies the computational ability of our theory. 
(3) We are dealing with quotient data, so we are actually manipulating
\textit{equivalence classes of terms} instead of one representative (term) from the equivalence class.
(4) Instead of using meta-level substitution, we use explicit
substitution and de Bruijn indices in our intrinsic formulation. This is
also called substitution calculus in the literature and is favoured due
to its closedness to categorical semantic. We will go through these four details one-by-one.

\textbf{Judgements}
We formalize our language as a deep embedding in a type theory with QIIT types. Every judgement is a (dependent) QIIT \textit{type} and all judgement derivations are well-typed QIIT terms. (Otherwise how can they be encoded in Agda?) There are four standard judgement forms in MLTT,
well-typed contexts $\goodCtx{\Gamma}{l}$, well-typed types
$\goodType{\Gamma}{T}{l}$, well-typed terms $\goodTerm{\Gamma}{t}{T}$, and
well-typed substitutions $\goodSub{\Gamma}{\gamma}{\Delta}$. These four judgements (QIIT types) are
mutually recursively defined together. Our presentation
follows the style of~\citet{kaposi2019gluing} and thus a predicative MLTT, but for readability we
omit the universe levels.
%\YZ{What's the distinction of judgments vs. sequents?}.\EDJreply{I am not sure about the correct terminology here. Here I want to say "even underivable judgement has to be well-typed". That's the main reason I spell out the (pseudo) Agda code there. I actually want the reader to have intrinsic-typed syntax in mind all the time (like read conventional math but acknowledge we are manipulating Agda)}\EDJreply{I rephrase that so there won't be confusion for the future readers. Please check}
Latex-wise, we spell out these \textit{well-formedness rules} of type
judgements and their (pseudo-)Agda counterparts to make things more accessible. Of
course, a well-formed judgement is not necessarily derivable, just as
a well-formed QIIT type is not necessarily inhabited. These well-formedness rules are direct translations of the \textit{well-typedness} of QIIT.

@@
